\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-02-27 papers/methods/paper.tex
Updating conclusion
Adding correlation section

@@

@@

In this paper we have outlined the creation of a hierarchical Bayesian model for supernova cosmology. The model takes into account selection effects and their uncertainty, fits underlying populations and standardisation parameters, incorporates unexplained dispersion from intrinsic scatter colour smearing and incorporates uncertainty from peculiar velocity uncertainty, calibration uncertainty, dust uncertainty and SALT2 model uncertainty. The model has been optimised to allow for hundreds of supernovae to be modelled fully with latent parameters and run in under an hour of CPU time and scales linearly with the number of supernovae, as opposed to polynomial complexity of matrix inversion.

>>>>

Table \ref{tab:parameter_correlations} list the fit correlations to our model fit parameters (excluding the LowZ band systematics, which had negligible correlation), and Figure \ref{fig:bulk_cor} show the full correlations between all non-systematic model parameters. As expected, $w$ bias is introduced primarily through incorrectly determined $\beta$, which itself is tied to the unexplained colour dispersion. Other interesting correlations are shown and discussed in Figure \ref{fig:bulk_cor}, however none have as significant an impact on cosmology than $\beta$. Table \ref{tab:parameter_correlations} does show that $\langle M_B \rangle$ is strongly correlated with $w$, however it is recovered much more accurately than $\beta$. The band systematics for DES filters $g$, $r$ and $i$ also show significant correlation with $w$, highlighting the importance of minimising instrumental uncertainty.

@@

\begin{table}
	\centering
	\caption{Parameter Correlations with $w$ for the combined 100 simulation fits. Correlations for the LowZ band systematics and the latent parameters representing selection function uncertainty are not shown but have negligible correlation.}
	\label{tab:parameter_correlations}
	\begin{tabular}{c|cc}
		Parameter & {\gten} Stat+Syst & {\celeven} Stat+Syst  \\
		\hline
		$\Omega_m$ &   -0.19 &     -0.21     \\
		$\alpha$ &   -0.17 &     -0.20     \\
		$\beta$ &   -0.29 &     -0.23     \\
		$\langle M_B \rangle$ &    0.68 &      0.66     \\
		$\sigma_{\rm m_B}^{0}$ &    0.04 &      0.07     \\
		$\sigma_{\rm m_B}^{1}$ &    0.23 &      0.18     \\
		$\sigma_{x_1}^{0}$ &    0.04 &      0.03     \\
		$\sigma_{x_1}^{1}$ &    0.05 &      0.01     \\
		$\sigma_{c}^{0}$ &    0.01 &      0.11     \\
		$\sigma_{c}^{1}$ &    0.08 &      0.04     \\
		$\alpha_c^{0}$ &   -0.04 &      0.04     \\
		$\alpha_c^{1}$ &    0.03 &      0.01     \\
		$\kappa_{c0}^{0}$ &   -0.10 &     -0.05     \\
		$\kappa_{c0}^{1}$ &   -0.20 &     -0.17     \\
		$\kappa_{c1}^{0}$ &   -0.05 &     -0.01     \\
		$\kappa_{c1}^{1}$ &   -0.01 &      0.01     \\
		$s_c^{0}$ &    0.00 &      0.01     \\
		$\delta(0)$ &    0.00 &      0.00     \\
		$\delta(\infty)/\delta(0)$ &    0.00 &      0.00     \\
		$\langle x_1^{0} \rangle$ &   -0.01 &     -0.05     \\
		$\langle x_1^{1} \rangle$ &   -0.02 &      0.02     \\
		$\langle x_1^{2} \rangle$ &   -0.04 &     -0.04     \\
		$\langle x_1^{3} \rangle$ &   -0.03 &     -0.06     \\
		$\langle x_1^{4} \rangle$ &   -0.06 &     -0.06     \\
		$\langle x_1^{5} \rangle$ &    0.04 &      0.02     \\
		$\langle x_1^{6} \rangle$ &    0.04 &      0.04     \\
		$\langle x_1^{7} \rangle$ &    0.08 &      0.03     \\
		$\langle c^{0} \rangle$ &   -0.05 &     -0.12     \\
		$\langle c^{1} \rangle$ &    0.11 &      0.03     \\
		$\langle c^{2} \rangle$ &    0.11 &      0.06     \\
		$\langle c^{3} \rangle$ &    0.14 &      0.04     \\
		$\langle c^{4} \rangle$ &   -0.11 &     -0.11     \\
		$\langle c^{5} \rangle$ &   -0.15 &     -0.08     \\
		$\langle c^{6} \rangle$ &   -0.12 &     -0.13     \\
		$\langle c^{7} \rangle$ &   -0.12 &     -0.06     \\
		$\delta [ {\rm SALT}_0 ]$ &    0.05 &      0.05     \\
		$\delta [ {\rm SALT}_1 ]$ &   -0.01 &      0.02     \\
		$\delta [ {\rm SALT}_2 ]$ &   -0.10 &     -0.09     \\
		$\delta [ {\rm SALT}_3 ]$ &   -0.03 &     -0.03     \\
		$\delta [ {\rm SALT}_4 ]$ &    0.08 &      0.09     \\
		$\delta [ {\rm SALT}_5 ]$ &    0.01 &      0.02     \\
		$\delta [ {\rm SALT}_6 ]$ &    0.05 &      0.07     \\
		$\delta [ {\rm SALT}_7 ]$ &   -0.11 &     -0.10     \\
		$\delta [ {\rm SALT}_8 ]$ &    0.01 &      0.02     \\
		$\delta [ {\rm SALT}_9 ]$ &    0.02 &      0.02     \\
		$\delta [ {\rm MWE}_{B-V} ]$ &    0.03 &      0.02     \\
		$\delta [ {\rm HST Calib} ]$ &   -0.07 &     -0.07     \\
		$\delta [ v_{\rm pec} ]$ &    0.00 &     -0.01     \\
		$\delta [ {\rm \delta z } ]$ &    0.01 &      0.00     \\
		$\delta [ \Delta g ]$ &    0.05 &      0.11     \\
		$\delta [ \Delta r ]$ &    0.16 &      0.10     \\
		$\delta [ \Delta i ]$ &   -0.16 &     -0.18     \\
		$\delta [ \Delta z ]$ &   -0.26 &     -0.26     \\
		$\delta [ \Delta \lambda_g ]$ &    0.16 &      0.20     \\
		$\delta [ \Delta \lambda_r ]$ &    0.05 &      0.06     \\
		$\delta [ \Delta \lambda_i ]$ &    0.00 &     -0.01     \\
		$\delta [ \Delta \lambda_z ]$ &    0.09 &      0.07     \\
		\hline
	\end{tabular}
\end{table}
%		$\delta [ \Delta t ]$ &   -0.01 &      0.00     \\
%		$\delta [ \Delta y ]$ &   -0.03 &     -0.04     \\
%		$\delta [ \Delta Z ]$ &    0.10 &      0.10     \\
%		$\delta [ \Delta A ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta u ]$ &   -0.04 &     -0.03     \\
%		$\delta [ \Delta v ]$ &    0.01 &      0.02     \\
%		$\delta [ \Delta \lambda_{t} ]$ &    0.00 &      0.02     \\
%		$\delta [ \Delta \lambda_{y} ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta \lambda_{z} ]$ &   -0.05 &     -0.04     \\
%		$\delta [ \Delta \lambda_{A} ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta \lambda_{u} ]$ &   -0.02 &     -0.02     \\
%		$\delta [ \Delta \lambda_{v} ]$ &   -0.01 &      0.01     \\
%		$\delta [ \Delta gl ]$ &   -0.12 &     -0.13     \\
%		$\delta [ \Delta hm ]$ &    0.02 &      0.01     \\
%		$\delta [ \Delta in ]$ &    0.08 &      0.13     \\
%		$\delta [ \Delta jo ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta \lambda_{gl} ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta \lambda_{hm} ]$ &   -0.01 &     -0.03     \\
%		$\delta [ \Delta \lambda_{in} ]$ &   -0.10 &     -0.12     \\
%		$\delta [ \Delta \lambda_{jo} ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta p ]$ &   -0.01 &     -0.01     \\
%		$\delta [ \Delta q ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta R ]$ &    0.00 &      0.01     \\
%		$\delta [ \Delta s ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta \lambda_{p} ]$ &    0.00 &      0.01     \\
%		$\delta [ \Delta \lambda_{q} ]$ &    0.01 &     -0.01     \\
%		$\delta [ \Delta \lambda_{r} ]$ &   -0.01 &     -0.01     \\
%		$\delta [ \Delta \lambda_{s} ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta b ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta c ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta d ]$ &    0.00 &      0.01     \\
%		$\delta [ \Delta e ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta \lambda_{b} ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta \lambda_{c} ]$ &    0.00 &      0.00     \\
%		$\delta [ \Delta \lambda_{d} ]$ &    0.00 &     -0.01     \\
%		$\delta [ \Delta \lambda_{e} ]$ &    0.00 &     -0.01     \\
%		$\Delta_{0}$ &    0.00 &      0.01     \\
%		$\Delta_{1}$ &    0.00 &     -0.01     \\
%		$\Delta_{2}$ &    0.00 &      0.00     \\
%		$\Delta_{3}$ &   -0.01 &     -0.01     \\
%		$\Delta_{4}$ &    0.00 &     -0.01     \\
%		$\Delta_{5}$ &    0.00 &     -0.01     \\
%		$\Delta_{6}$ &    0.00 &      0.00     \\
%		$\Delta_{7}$ &    0.00 &     -0.01     \\

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{bulk_cor.pdf}
	\end{center}
	\caption{Parameter correlations for the combined fits to the 100 {\gten} scatter model simulations. We see that the primary correlations with $w$ enter through $\alpha$, $\beta$ and $\langle M_B \rangle$, as shown in Table \ref{tab:parameter_correlations}. Also visible in this figure are several other interesting relationships. $\beta$ is strongly anti-correlated with intrinsic dispersion $\sigma_{m_B}$ for both surveys (DES like and LowZ), with $\sigma_{m_B}$ showing strong anti-correlation with $\kappa^0_c$. This relationship is indeed expected - as $\kappa^0_c$ grows larger (more unexplained dispersion on the colour observation), the width of the supernova population in apparent magnitude space increases. As the fit prefers it to conform to the observed width of the distribution, the extra width in colour causes the inherent magnitude smearing amount to decrease. And with extra freedom on the observed colour from $\kappa^0_c$, $\beta$ shifts in response. The other striking feature in the plot are the strong correlation blocks in the bottom right and the anti-correlation stripes on the edges. These too are expected, for they show the relationship between the colour distributions mean value, its width and its skewness. As skewness or population width increases, the effective mean of the population shifts (see Appendix \ref{app:approx} for details), creating anti-correlation between skewness and the (Gaussian) mean colour population. Finally, we notice the blank cross in the middle of the figure, the result of four parameters. Firstly, the redshift-dependent extra colour dispersion for the LowZ simulation is negligible (due to the low redshift values), the bias correciton $s_c^0$ is unconstrained by our fits, and as the SNANA simulations do not include mass, we expect zero correlation across the mass step $\delta(0)$ and $\delta(\infty)/\delta(0)$ parameters. }
	\label{fig:bulk_cor}
\end{figure*}

@@

In this paper we have outlined the creation of a hierarchical Bayesian model for supernova cosmology. The model takes into account selection effects and their uncertainty, fits underlying populations and standardisation parameters, incorporates unexplained dispersion from intrinsic scatter colour smearing and incorporates uncertainty from peculiar velocity uncertainty, calibration uncertainty, dust uncertainty and SALT2 model uncertainty. Furthermore, these uncertainties, being explicitly parametrised in our model, are captured will covariance intact, an improvement on many previous methods. The model has been optimised to allow for hundreds of supernovae to be modelled fully with latent parameters and run in under an hour of CPU time and scales linearly with the number of supernovae, as opposed to polynomial complexity of matrix inversion.
